A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-20 arxiv/setup.tex
wip

Following \citet{HMRT19} and \citet{BLLT20},
%In the high-dimensional linear regression setting (e.g. ), the features of the $k$-th task, denoted by $X_k\in\real^{n_i\times p}$, consist of $n_k$ feature vectors given by $x_1, x_2, \dots, x_{n_k}$.
we assume that for each task $k$ from $1$ to $t$, each feature vector $x_i = \Sigma^{1/2}_k z_i$ for $i = 1, \dots, n_k$, where $z_i\in\real^p$ is an i.i.d. random vector with  mean zero and unit variance.
The sample size of task $k$, given by $n_k$, is equal to $\rho_k\cdot p$ for a fixed value $\rho_k$ that does not grow with feature dimension $p$.
This setting is also known as the high-dimensional linear regression setting in the literature.
There are two motivations for studying the high-dimensional linear regression setting for multi-task learning.
First, this setting captures salient properties of modern large-scale datasets, where the sample sizes are usually on the order of tens to hundreds of the number of features \cite{sur2019modern}.
Second, as we will see soon in Section \ref{sec_main}, we can derive precise asymptotics of the generalization error that scale with properties of the task data such as sample sizes.
%The labels $Y_k = X_k \beta_k + \varepsilon_k$, where $\beta_k$ denotes the linear model parameters and $\varepsilon_k$ denotes i.i.d. noise with mean zero and variance $\sigma^2$.
%Recall that we have $t$ labeled training datasets, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$, where $X_i\in\real^{n_i\times p}$ and $Y_i\in\real^{n_i}$ for $1\le i\le t$.
%Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
%Without loss of generality, let the $t$-th task be the target task.

\medskip
\noindent\textbf{Hard parameter sharing.}
We study multi-task learning models that use hard parameter sharing, which is the most commonly used architecture in practice \cite{R17}.
Specifically, we consider a two-layer neural network with a shared layer $B\in\real^{p\times r}$ for all tasks and $t$ output layers $W_1, \dots, W_t$ for every task from $1$ to $t$, where each output layer is a $r$-dimensional vector.
The width of $B$, denoted by $r$, plays an important role in regularization.
As observed in Proposition 1 of \cite{WZR20}, if $r \ge t$, there is no regularization effect.
Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following procedure inspired by how hard parameter sharing models are trained in practice.
\begin{enumerate}
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
		{\begin{align}\label{eq_mtl}
			f(B; W_1, \dots, W_t) = \sum_{k=1}^t \norm{X_k B W_k - Y_k}^2.
		\end{align}}
	\item Learn the output layers $W_1, W_2, \dots, W_t$: set $B = \hat{B}$ and minimize the training loss over $W_1, W_2, \dots, W_t$.
		{\begin{align}\label{eq_mtl_eval}
			g(W_1, \dots, W_t) = \sum_{k=1}^t \norm{X_k \hat{B} W_k - Y_k}^2.
		\end{align}}
\end{enumerate}
Let $\hat{\beta}_t^{\MTL}$ denote the multi-task learning estimator obtained from the procedure above.
By contrast, let $\hat{\beta}_t^{\STL} = (X_t^{\top}X_t)^{-1}X_t^{\top}{Y_t}$ denote the single-task learning estimator. % denoted by $L(\hat{\beta}_t^{\STL})$.
%where $B\in\real^{p\times r}$ and $W_k\in\real^r$ for every $1\le k\le t$.
%Following , we assume that $r < t$, because otherwise minimizing $f(\cdot)$ could result in $BW_i$ being the single-task optimum.
In general, the objective $f(\cdot)$ is non-convex in $B$ and the $W_k$'s.
Therefore, we first minimize $B$ in equation \eqref{eq_mtl} and then minimize $W_k$ given $B$ in equation \eqref{eq_mtl_eval}.
%For our purpose, a validation set that is much larger than the number of output layer parameters $r\cdot t$ suffices (e.g. $\rho_i \cdot p^{0.5}$).
%The size of the training set is then $\rho_i (p - p^{0.99})$.
%The advantage of tuning the output layers on the validation set is to reduce the effect of noise from $\hat{B}$.
%For more details, we refer the reader to Appendix \ref{app_proof_sec3}.

%The single-task estimator $\hat{\beta}_t^{\STL}$ is given by $(X_t^{\top}X_t)^{-1}X_t^{\top}Y_t$.
%For an estimator $\hat{\beta}\in\real^p$, we define the out-of-sample prediction loss as

%which can be further decomposed as the bias plus the variance of $\hat{\beta}$.
%In order to relate the btradeoff to properties of the data, we need tight concentration bounds for the bias and variance.
%We consider the high-dimensional regime where $n_i$ is a fixed constant $\rho_i > 1$  multiple of $p$ for every $1\le i\le t$, and $p$ is sufficiently large.
%We focus on a setting where $\rho_t$ is small compared to $\set{\rho_i}_{i=1}^{t-1}$.
%This setting captures the need to add more labeled data to reduce the prediction loss of the target task.

%However, this result only applies to a single task.
%Therefore, our goal is to extend this result to multiple tasks.

%\textbf{Notations.}
%When there is no ambiguity, we drop the subscript $t$ from $\te_t(\hat{\beta}_t^{\MTL})$ and write $\te(\hat{\beta}_t^{\MTL})$ for simplicity.
%We refer to the first task as the source task when there are only two tasks.
%We call $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ the covariate shift matrix.

%\subsection{Analyzing the Tradeoff via Random Matrix Theory}

\medskip
\noindent\textbf{Problem statement.}
%We focus on predicting a particular task, say the $t$-th task, without loss of generality.
For an estimator $\hat{\beta}$ of the target task model $\beta_t$, we define the prediction loss of the estimator $\hat{\beta}$ as
	{\begin{align*}
		\te(\hat{\beta}) = \exarg{x = \Sigma_t^{1/2} z}{({x}^{\top}\hat{\beta} - {x}^{\top}\beta_t)^2}
		= \bignorm{\Sigma_t^{1/2} (\hat{\beta} - \beta_t)}^2,
	\end{align*}}%
where $x = \Sigma_t^{1/2} z$ denotes a random feature vector with covariance $\Sigma_t$.
In the above equation, $x^{\top}\beta_t$ is the true label of $x$.
%= (\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t)
We say that the source tasks provide a \textit{positive transfer} to the target task if the prediction loss of the MTL estimator is lower than that of the STL estimator, that is, if
	\[ L(\hat{\beta}_t^{\MTL}) > L(\hat{\beta}_t^{\STL}). \]
On the other hand, we say that the source tasks provide a \textit{negative transfer} to the target task if $L(\hat{\beta}_t^{\MTL}) < L(\hat{\beta}_t^{\STL})$.
Our goal is to study when the source tasks provide a positive transfer to the target task.

>>>>

Following \citet{HMRT19} and \citet{BLLT20},
%In the high-dimensional linear regression setting (e.g. ), the features of the $k$-th task, denoted by $X_k\in\real^{n_i\times p}$, consist of $n_k$ feature vectors given by $x_1, x_2, \dots, x_{n_k}$.
we assume that for each task $k$ from $1$ to $t$, each feature vector $x_i = \Sigma^{1/2}_k z_i$ for $i = 1, \dots, n_k$, where $z_i\in\real^p$ is an i.i.d. random vector with  mean zero and unit variance.
The sample size of task $k$, given by $n_k$, is equal to $\rho_k\cdot p$ for a fixed value $\rho_k$ that does not grow with feature dimension $p$.
This setting is also known as the high-dimensional linear regression setting in the literature.
There are two motivations for studying the high-dimensional linear regression setting for multi-task learning.
First, this setting captures salient properties of modern large-scale datasets, where the sample sizes are usually on the order of tens to hundreds of the number of features \cite{sur2019modern}.
Second, as we will see soon in Section \ref{sec_general}, we can derive precise asymptotics of the generalization error that scale with properties of the task data such as sample sizes.
%The labels $Y_k = X_k \beta_k + \varepsilon_k$, where $\beta_k$ denotes the linear model parameters and $\varepsilon_k$ denotes i.i.d. noise with mean zero and variance $\sigma^2$.
%Recall that we have $t$ labeled training datasets, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$, where $X_i\in\real^{n_i\times p}$ and $Y_i\in\real^{n_i}$ for $1\le i\le t$.
%Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
%Without loss of generality, let the $t$-th task be the target task.

\medskip
\noindent\textbf{Hard parameter sharing.}
We study multi-task learning models that use hard parameter sharing, which is the most commonly used architecture in practice \cite{R17}.
Specifically, we consider a two-layer neural network with a shared layer $B\in\real^{p\times r}$ for all tasks and $t$ output layers $W_1, \dots, W_t$ for every task from $1$ to $t$, where each output layer is a $r$-dimensional vector.
The width of $B$, denoted by $r$, plays an important role in regularization.
As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following procedure inspired by how hard parameter sharing models are trained in practice.
\begin{enumerate}
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
		{\begin{align}\label{eq_mtl}
			f(B; W_1, \dots, W_t) = \sum_{k=1}^t \norm{X_k B W_k - Y_k}^2.
		\end{align}}
	\item Learn the output layers $W_1, W_2, \dots, W_t$: set $B = \hat{B}$ and minimize the training loss over $W_1, W_2, \dots, W_t$.
		{\begin{align}\label{eq_mtl_eval}
			g(W_1, \dots, W_t) = \sum_{k=1}^t \norm{X_k \hat{B} W_k - Y_k}^2.
		\end{align}}
\end{enumerate}
Let $\hat{\beta}_t^{\MTL}$ denote the multi-task learning estimator obtained from the procedure above.
By contrast, let $\hat{\beta}_t^{\STL} = (X_t^{\top}X_t)^{-1}X_t^{\top}{Y_t}$ denote the single-task learning estimator. % denoted by $L(\hat{\beta}_t^{\STL})$.
%where $B\in\real^{p\times r}$ and $W_k\in\real^r$ for every $1\le k\le t$.
%Following , we assume that $r < t$, because otherwise minimizing $f(\cdot)$ could result in $BW_i$ being the single-task optimum.

\smallskip
\noindent\textit{Remark.}
In general, the multi-task learning objective $f(\cdot)$ is non-convex with respect to $B$ and $W_1, \dots, W_t$.
Therefore, we first minimize $B$ in equation \eqref{eq_mtl} and then minimize $W_k$ given $B$ in equation \eqref{eq_mtl_eval}.
For our results later in Section \ref{sec_general} and \ref{sec_special}, we will identify tractable cases and provide guarantees to the above procedure.

\medskip
\noindent\textbf{Problem statement.}
%We focus on predicting a particular task, say the $t$-th task, without loss of generality.
For an estimator $\hat{\beta}$ of the target task model $\beta_t$, we define the prediction loss of the estimator $\hat{\beta}$ as
	{\begin{align*}
		\te(\hat{\beta}) = \exarg{x = \Sigma_t^{1/2} z}{({x}^{\top}\hat{\beta} - {x}^{\top}\beta_t)^2}
		= \bignorm{\Sigma_t^{1/2} (\hat{\beta} - \beta_t)}^2,
	\end{align*}}%
where $x = \Sigma_t^{1/2} z$ denotes a random feature vector with covariance $\Sigma_t$.
In the above equation, $x^{\top}\beta_t$ is the true label of $x$.
%= (\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t)
We say that the source tasks provide a \textit{positive transfer} to the target task if the prediction loss of the MTL estimator is lower than that of the STL estimator, that is, if
	\[ L(\hat{\beta}_t^{\MTL}) > L(\hat{\beta}_t^{\STL}). \]
On the other hand, we say that the source tasks provide a \textit{negative transfer} to the target task if $L(\hat{\beta}_t^{\MTL}) < L(\hat{\beta}_t^{\STL})$.
Our goal is to study when the source tasks provide a positive transfer to the target task.
More specifically, we study how varying properties of task data including task similarity, sample ratio, and covariate shift affects information transfer in multi-task learning.

### 2020-08-20 arxiv/transfer.tex
wip

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
We show theoretical implications for understanding the performance of multi-task learning.
(a) \textit{Task similarity}: we explain the phenomenon of negative transfer precisely as tasks become more different.
(b) \textit{Sample size}: we further explain a curious phenomenon where increasing the source sample size helps initially, but hurts eventually.
(c) \textit{Covariate shift}: as the source sample size increases, we show that the covariate shift worsens the performance of the multi-task estimator.
Finally, we extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.

@@

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Roughly speaking, the transition threshold scales as $\frac{pd^2}{\sigma^2} - \frac{1}{\rho_1} - \frac{1}{\rho_2}$.
We apply Proposition \ref{prop_dist_transition} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}).
We can see that our result is able to predict positive or negative transfer  accurately and matches the empirical curve.
There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}.
A key part of the analysis shows that $\hat{v}\approx 1$ in the isotropic model,
thus simplifying the result of Theorem \ref{thm_main_informal}.
%, which is based on our main result described later in Theorem \ref{thm_model_shift}.

@@

Next we consider the case where the two tasks have different noise variances $\sigma_1^2\ne \sigma_2^2$. In particular, we show Proposition \ref{prop_var_transition}, which gives a transition threshold with respect to the difference between the noise levels of the two tasks.

\begin{proposition}%[Labeled data de-noising]
\label{prop_var_transition}
	In the isotropic model, assume that $\rho_1 > 40$ and $\ex{\norm{\beta_1 - \beta_2}^2}< \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$.
	Then we have the following transition with respect to $\sigma_1^2$:
	\begin{itemize}
		\item If $\sigma_1^2 < - \nu^{1/2} \rho_1 \cdot p d^2+\left(1+ \nu^{-1/2}\rho_1 \Phi(\rho_1, \rho_2)\right)\cdot\sigma_2^2$, then w.h.p. $\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL})$.
		\item If $\sigma_1^2 > -\nu^{-1/2}\rho_1\cdot p d^2   +\left(1+ \nu^{1/2}\rho_1\Phi(\rho_1, \rho_2)\right) \cdot \sigma_2^2$, then w.h.p. $\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL})$.
	\end{itemize}
\end{proposition}
As a corollary, if $\sigma_1^2 \le \sigma_2^2$, then we always get positive transfer.

\textbf{Algorithmic consequence.}
We can in fact extend the result to the cases where the noise variances are different.
In this case, we will see that MTL is particularly effective.
%As an extension of Proposition \ref{prop_dist_transition}, we observe that MTL is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Concretely, suppose the noise variance $\sigma_1^2$ of task $1$ differs from the noise variance $\sigma_2^2$ of task $2$.
If $\sigma_1^2$ is too large, the source task provides a negative transfer to the target.
If $\sigma_1^2$ is small, the source task is more helpful.
We leave the result to Proposition \ref{prop_var_transition} in Appendix \ref{app_proof_31}.
Inspired by the observation, we propose a single-task based metric to help understand MTL results using STL results.
\squishlist
	\item For each task, we train a single-task model.
	Let $z_s$ and $z_t$ be the prediction accuracy of each task, respectively.
	Let $\tau\in(0, 1)$ be a fixed threshold.
	\item If $z_s - z_t > \tau$, then we predict that there will be positive transfer when combining the two tasks using MTL.
	If $z_s - z_t < -\tau$, then we predict negative transfer.
\squishend

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}%
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.

@@

\begin{proposition}%[Labeled data efficiency]
\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1,\rho_2 \ge 9$ and $\Psi(\beta_1, \beta_2) < (5(\rho_1-1))^{-1} + (5(\rho_2-1))^{-1}$.
	Then the data efficiency ratio $x^\star$ satisfies
	\be\label{eq_uplowx} x_l  \le x^\star\le \frac{1}{\rho_1 + \rho_2} \bigbrace{  \frac{2}{(\rho_1-1)^{-1} + (\rho_2-1)^{-1} - 5\Psi(\beta_1, \beta_2)}+1}, \ee
	where we denoted
	$$x_l:= \frac1{\rho_1+\rho_2}\left(\frac{2}{(\rho_1-1)^{-1}+(\rho_2 -1)^{-1}}+1\right).$$
\end{proposition}

\textbf{Algorithmic consequence.} An interesting consequence of Proposition \ref{prop_data_size} is that $L(\hat{\beta}_t^{\MTL})$ is not monotone in $\rho_1$.
In particular, Figure \ref{fig_size} (and our analysis) shows that $L(\hat{\beta}_t^{\MTL})$ behaves as a quadratic function over $\rho_1$.
More generally, depending on how large $\Psi(\beta_1, \beta_2)$ is, $L(\hat{\beta}_t^{\MTL})$ may also be monotonically increasing or decreasing.
Based on this insight, we propose an incremental optimization schedule to improve MTL training efficiency.
\squishlist
	\item We divide the source task data into $S$ batches.
	For $S$ rounds, we incrementally add the source task data by adding one batch at a time.
	\item After training $T$ epochs, if the validation accuracy becomes worse than the previous round's result, we terminate.
	Algorithm \ref{alg_inc_train} in Appendix \ref{app_experiments} describes the procedure in detail.
\squishend

@@

\textbf{Algorithmic consequence.}
Our observation highlights the need to correct covariate shift when $n_1 / n_2$ is large.
Hence for such settings, we expect procedures that aim at correcting covariate shift to provide more significant gains.
We consider a covariance alignment procedure proposed in \cite{WZR20}, which is designed for the purpose of correcting covariate shift.
The idea is to add an alignment module between the input and the shared module $B$.
This new module is then trained together with $B$ and the output layers.
We validate our insight on this procedure in the experiments.

>>>>

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
We show theoretical implications for understanding the performance of multi-task learning.
(a) \textit{Task similarity}: we explain the phenomenon of negative transfer precisely as tasks become more different.
(b) \textit{Sample ratio}: we further explain a curious phenomenon where increasing the sample ratio helps initially, but hurts eventually.
(c) \textit{Covariate shift}: as the source sample size increases, we show that the covariate shift worsens the performance of the multi-task estimator.
Finally, we extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.

@@

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Roughly speaking, the transition threshold scales as $\frac{pd^2}{\sigma^2} - \frac{1}{\rho_1} - \frac{1}{\rho_2}$.
We apply Proposition \ref{prop_dist_transition} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}).
We can see that our result is able to predict positive or negative transfer  accurately and matches the empirical curve.
There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}.
A key part of the analysis shows that $\hat{v}\approx 1$ in the isotropic model,
thus simplifying the result of Theorem \ref{thm_main_informal}.

@@

\paragraph{Benefits of multi-task learning.}
We can in fact extend the result to the cases where the noise variances are different.
In this case, we will see that MTL is particularly effective.
%As an extension of Proposition \ref{prop_dist_transition}, we observe that MTL is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Concretely, suppose the noise variance $\sigma_1^2$ of task $1$ differs from the noise variance $\sigma_2^2$ of task $2$.
If $\sigma_1^2$ is too large, the source task provides a negative transfer to the target.
If $\sigma_1^2$ is small, the source task is more helpful.
We leave the result to Proposition \ref{prop_var_transition} in Appendix \ref{app_proof_31}.
Next we consider the case where the two tasks have different noise variances $\sigma_1^2\ne \sigma_2^2$. In particular, we show Proposition \ref{prop_var_transition}, which gives a transition threshold with respect to the difference between the noise levels of the two tasks.

\begin{proposition}[Label denoising]
\label{prop_var_transition}
	In the isotropic model, assume that $\rho_1 > 40$ and $\ex{\norm{\beta_1 - \beta_2}^2}< \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$.
	Then we have the following transition with respect to $\sigma_1^2$:
	\begin{itemize}
		\item If $\sigma_1^2 < - \nu^{1/2} \rho_1 \cdot p d^2+\left(1+ \nu^{-1/2}\rho_1 \Phi(\rho_1, \rho_2)\right)\cdot\sigma_2^2$, then w.h.p. $\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL})$.
		\item If $\sigma_1^2 > -\nu^{-1/2}\rho_1\cdot p d^2   +\left(1+ \nu^{1/2}\rho_1\Phi(\rho_1, \rho_2)\right) \cdot \sigma_2^2$, then w.h.p. $\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL})$.
	\end{itemize}
\end{proposition}
As a corollary, if $\sigma_1^2 \le \sigma_2^2$, then we always get positive transfer.

We use our tools to explain a key result of Taskonomy by Zamir et al. \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}%
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.

@@

\begin{proposition}[Labeling efficiency]
\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1,\rho_2 \ge 9$ and $\Psi(\beta_1, \beta_2) < (5(\rho_1-1))^{-1} + (5(\rho_2-1))^{-1}$.
	Then the data efficiency ratio $x^\star$ satisfies
	\be\label{eq_uplowx} x_l  \le x^\star\le \frac{1}{\rho_1 + \rho_2} \bigbrace{  \frac{2}{(\rho_1-1)^{-1} + (\rho_2-1)^{-1} - 5\Psi(\beta_1, \beta_2)}+1}, \ee
	where we denoted
	$$x_l:= \frac1{\rho_1+\rho_2}\left(\frac{2}{(\rho_1-1)^{-1}+(\rho_2 -1)^{-1}}+1\right).$$
\end{proposition}

\paragraph{Algorithmic consequence.}

\noindent\textit{Detecting negative transfer.}
Inspired by the observation, we propose a single-task based metric to help understand MTL results using STL results.
\squishlist
	\item For each task, we train a single-task model.
	Let $z_s$ and $z_t$ be the prediction accuracy of each task, respectively.
	Let $\tau\in(0, 1)$ be a fixed threshold.
	\item If $z_s - z_t > \tau$, then we predict that there will be positive transfer when combining the two tasks using MTL.
	If $z_s - z_t < -\tau$, then we predict negative transfer.
\squishend

\noindent\textit{Mitigating negative transfer.}
An interesting consequence of Proposition \ref{prop_data_size} is that $L(\hat{\beta}_t^{\MTL})$ is not monotone in $\rho_1$.
In particular, Figure \ref{fig_size} (and our analysis) shows that $L(\hat{\beta}_t^{\MTL})$ behaves as a quadratic function over $\rho_1$.
More generally, depending on how large $\Psi(\beta_1, \beta_2)$ is, $L(\hat{\beta}_t^{\MTL})$ may also be monotonically increasing or decreasing.
Based on this insight, we propose an incremental optimization schedule to improve MTL training efficiency.
\squishlist
	\item We divide the source task data into $S$ batches.
	For $S$ rounds, we incrementally add the source task data by adding one batch at a time.
	\item After training $T$ epochs, if the validation accuracy becomes worse than the previous round's result, we terminate.
	Algorithm \ref{alg_inc_train} in Appendix \ref{app_experiments} describes the procedure in detail.
\squishend

@@

\noindent\textit{Covariance alignemnt.}
Our observation highlights the need to correct covariate shift when $n_1 / n_2$ is large.
Hence for such settings, we expect procedures that aim at correcting covariate shift to provide more significant gains.
We consider a covariance alignment procedure proposed in \cite{WZR20}, which is designed for the purpose of correcting covariate shift.
The idea is to add an alignment module between the input and the shared module $B$.
This new module is then trained together with $B$ and the output layers.
We validate our insight on this procedure in the experiments.
